### 2009-12-11 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Revision of Prof. Puder's audio feedback of chapter 5, as well as entire restructure of the document, such as using more page width and height, making it more visual. Table references not working automatically, as well as page numbers (will review later). Table references are entered manually in the text.
The structure of chapter 5 changed according to the fact that some
sections, subsections were very short, with 1 paragraph. So, I decided
to use two different options:

* If there are more to be discussed, a new advantages/disadvantages
subsection is created; or else, one section includes two

%\bibliography{thesis}  % uses the references stored in Chapter1Radar.bib

\chapter{Empirical Analysis of Data Persistence Taxonomies for NetBEAMS}

@@

\begin{itemize}
  \item Provide data management over the collected sensor data, as a
  conventional database systems;
  \item Must be able to scale in any increases to the data load of
  the sensor network, as described in chapter 2;
  \item  Must be able to use single or multiple servers to improve performance
  and optionally deal with Data-Centric approaches to sensor data partitioning;
  \item Must provide a data model that scales, decreasing the constant data
  schema changes.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  data warehouse with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in Section 
  \ref{sec:sn-persitence-storage}, TinyDB is a database system used in many
  different sensor networks implementation such as \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source KVP database
  system suggested by different surveys to be alternatives to relational
  databases \cite{db-is-rdbs-dommed, cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. Furthermore, it was selected because
  NetBEAMS represents serialized objects in XML, as shown
  \cite{netbeams-dsp-architecture}.
\end{itemize}

The following sections review these technologies' capabilities to adhere to
the specifications of the taxonomies defined in chapter 3, as well as the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

As shown in the previous chapter, NetBEAMS's architecture is based on sensor
network whose topology is a single-hop star with a centralized data sink. In
this way, all the collected data is transmitted to that single location
\ref{sec:sn-infrastructure}. The simplest location to save the collected data
is through the use of at least an \textbf{External Storage}. Similarly, the
use of a Data-Centric Storage strategy is considered in situations where the
data sink is about to reach its storage capacity. The \textbf{Data-Centric
Storage} separates the collected data into dissimilar locations, based on any
property of the data. Therefore, there are two different approaches to data
partitioning:

\begin{itemize}
  \item \textbf{Horizontal Partitioning}: the data segmentation is implemented
  based on the values of the table rows. That is, all the values defined by the
  columns are used in a given table physically located in specific servers.
  For instance, horizontal partitioning can be defined as a collection of five
  years worth of historical sensor data, partitioned into five separate
  servers. In this way, each server contains data related to each of the years;
  \item \textbf{Vertical Partitioning}: data is partitioned vertically in a way
  that the columns of an entity are divided into different database tables. In
  this way, the particular dataset is broken down to different tables. For
  example, if the database is used to store pictures of a given sensor camera
  in a BLOB column, it would be the most preferable to the scenario where the
  images are not constantly referenced when compared to the textual data.
\end{itemize}

\subsection{Advantages of the External Storage Approach}

The External Storage approach is the most common way to implement persistence
for any type of systems since data management is usually offered by most
database systems out-of-the-box.

\subsection{Disadvantages of the External Storage Approach}

The most common problems related to a single External Data Storage approach are
related to the capacity of the device and operational problems. If the
external storage device is situated in the network, then the capacity of the
device may be limited to receive collected data from the sensors in the
network, since it may run out of space. Similarly, storage devices are very
error prone and its hardware part may fail.

@@

\begin{itemize}
  \item \textbf{Hash Partitioning}: a hash key is computed based on
  different table columns of an entity \cite{db-shard-schemas,
  db-partitioning-relational}, providing an even distribution of the data;
  \item \textbf{Key Partitioning}: is based on the hash partitioning over the
  values of a specific key \cite{db-mongo-partition};
  \item \textbf{Range Partitioning}: the partition is based on specific ranges
  of the values of given a given column of a table. For instance, the collected
  data from the sensors could be separated by the year of its collection. For
  example, data collected in the year of 2009 would be stored in a different
  partition than the data collected in the year 2010
  \cite{db-partitioning-relational};
  \item \textbf{Directory Based Partitioning}: an ordinary lookup service is
  used to identify which database partition to use \cite{db-shard-schemas}.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/database-sharding-by-region}
  \caption{Example of Data-Centric Storage for Sensor Networks using Database
  Sharding}
  \label{fig:database-sharding-by-region}
\end{figure}

@@

\begin{itemize}
  \item \textbf{Rebalancing Partitions}: considering a schema-dependent database
  system, changes on the schema of one shard requires rebalancing the
  collected data on each of the shards. As a consequence, the database shards
  transfers data to new locations whenever necessary. This technique is
  starting to appear as COTS implementations such as MySQL \cite{mysql} and
  mongoDB \cite{mongodb};
  \item \textbf{Referential integrity}: also considering a schema-dependent
  database system, any cross-shard query may run into the problem where
  references do no exist in other shards, since the application layer is
  responsible for enforcing data integrity. In this way, examples of such
  include verifying constraints of foreign keys when the partition is done
  using Vertical table partitioning, considering that data collections are
  organized in separate spaces, and physically placed in different shard nodes
  \cite{db-partitioning-relational}.
\end{itemize}

@@

One of the most common practices in the area of database system is to use the
relational model for data persistence, although the application of the system
may not fit to solve the problem. In fact, the main users of sensor networks may
not possess any expertise in database systems or data modeling, based on the
fact that they are from different science areas. \cite{sn-programming-language}
suggested the use of programming languages abstraction, instead of regular
database systems, to provide access to the collected data for both
technical and non-technical users. For instance, marine biologists manage
SF-BEAMS without expertise in data management or modeling systems. For this
reason, one of the requirements of the system is the use of schema-less
approach, which describes the collected data without data modeling process
used in schema-dependent models. The \textbf{Key-Value-Pair Data Model} seem
to be an optional candidate that provides such a method of data design.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/Relational-Model-Original}
  \caption{Relational Data Model for NetBEAMS - A first prototype}
  \label{fig:Relational-Model-Original}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/Relational-Model-Addition-Modified}
  \caption{Relational Data Model for NetBEAMS - Modified version with new
  entity}
  \label{fig:Relational-Model-Addition-Modified}
\end{figure}

\begin{itemize}
  \item Data schema that constantly change requires continuous database
  normalization processes, changes to structure, database management, and all
  the code that uses the data. Schema-Dependent databases always require
  refactoring of the schema to accommodate changes to the entities' properties;
  \item \cite{sn-provenance} describes the relational model as problematic 
  when describing collected data from sensor devices;
  \item \cite{sn-db-newop} suggests changes to the SQL clauses of TinyDB to
  better support time-series, an important property when describing.
\end{itemize}

Other projects have used the XML data models for persistence. It falls
into the same category as the Relational Model using XML documents \cite{xml},
since XML documents must fulfill the specifications defined by an XML Schema
\cite{xml-schema}. This model uses XPath \cite{xml-xpath} technology for
querying documents, although some hybrid \cite{db2} technologies may still use
SQL \cite{sql} for that matter. In this way, a materialized version of the
data from the defined previous schema can be seen in Figure
\ref{fig:persistence-example-relational}.

@@

Although this data model has been used in different types of applications,
developers have tried to use the Relational Model to design an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. As mentioned earlier, with the
emergence of Internet applications, the necessity of such a model that does
not require constant refactoring led developers to propose the relational model
as documented in technical articles such as \cite{db-kvp-in-relational01} and
\cite{db-kvp-in-relational02}. Based on these articles, consider the
relational model depicted in Figure \ref{fig:KVP-on-Relational-Model} as a
data model for persisting data to our case study using the key-value
pair strategy. It is clear that this approach contains problems: repetition
occurs, since the nature of time-series data requires the use of a timestamp
key for each property of the sensor. Therefore, this strategy is not well
suited to provide persistence data from collected sensor networks.

@@

The growth of distributed systems and the Internet have enabled the development
of powerful database servers that can be organized in the context of
infrastructure and how to represent models. A powerful up-coming approach is
the use of database systems that do not use structured language for querying
its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, which is also known as Document-Oriented, Distributed Hash
Table. Such a model has the following properties:

@@

There are no records of the use of this data model with Sensor Networks.
A different variant of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure, freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json}. mongoDB implements the
following:

\begin{itemize}
  \item entities from the same domains are placed into a ``bucket'';
  \item entities have a set of attributes and relating values
  \item entities with different set of attributes may be contained in the same
  bucket, since there is no schema to govern the bucket items restriction.
\end{itemize}

For instance, all data needed for the YSI Sonde Data, as well as all necessary
provenance data that describes the data depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and is compared on Table
\ref{tab:schema-vs-schemaless}:

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model is defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships \cite{db-normalization};
      \item Data Provenance can be provided through data types such as
      ``timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enables them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:schema-vs-schemaless-daccess}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic CRUD (Create-Retrieve-Update-Delete)
      database operations are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, containing aggregation functions to apply on the returned values.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via an API (Application
      Programming Interface) through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like equals to (=), different than (!=), less than
      (<), greater than (>), among others;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:tab:schema-vs-schemaless-api-interface}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC (Open Database
      Connectivity);
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Tend to provide Web Services Interfaces such as the REST
      \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

The following summarizes which data model each of the candidate database
systems use.

\begin{itemize}
  \item \textbf{MySQL}: uses the Relational Data Model with the use of SQL;
  \item \textbf{TinyDB}: uses the Relational Data Mode with the use of SQL; 
  \item \textbf{MongoDB}: uses Key-Value Pair Data Model with an Non-Structure
  Query language in forms of programming language abstraction;
  \item \textbf{IBM DB2}: uses the Relational and Structured models with SQL
  and XPath, respectively.
\end{itemize}

@@

\subsection{Advantages of the Provenance-Aware Data}

The inclusion of provenance data into the collected data enriches the
description of a given event, and is always useful to determine different
aspects of the data. As mentioned in chapter 3, data provenance helps identify
the nature of the data by the identification of the properties of a given
sensor device, for example. Similarly, provenance adds spatial-time data in
order to correlate the observed data with time a given location and time of
occurrence.

\subsection{Disadvantages of the Provenance-Aware Data}

Additional data being inserted together with the raw data collected from the
sensor devices may contribute with a decreasing amount of disk storage. 
Therefore, the use of provenance-away data increases the required disk space
used to save the collected data from sensor devices.

\subsection{Advantages of the Annotations}

Annotations are used to describe the collect data even more, since the
additional ``notes'' of the data are being stored together with the data. For
example, the use of tags to a given entity may help identifying and different
observations regarding the data collected from sensor devices.

\subsection{Disadvantages of the Annotations}

This type of data does not make up part of the collected data, nor provenance
information.

@@

All the database systems technology supports the use of Data Provenance or
Annotations. The Key-Value Pair data model has the advantage of not maintaining
a schema design, and therefore, can add any type of data, and therefore,
mongoDB is the lead in this taxonomy.

@@

\subsection{Advantages of the Centralized Query Processing}

Centralized data management and query processing is simpler than the in-network
query method. Data is verified by a straightforward database management system
without the risk of data being unavailable, when the data is spread among the
network nodes.

\subsection{Disadvantages of the Centralized Query Processing}

When the Centralized Query processing is used, problems may occur. Depending on
the size of the datasets and the database technology, performance may be a
concern due to the alleged Funneling Affect \cite{sn-storage04}, since the
point-of-traffic is concentrated in the centralized system.

@@

Sensor Network data can be saved into database system that are organized
by either a Centralized or Distributed System. While a centralized database
system may be considered to be easier to manage, it may fail to adjust to
changes of requirements to manage data. On the other hand, a distribute
database system can be used to implement a data-centric solution. Therefore, a
database system that can be organized in both approaches is a good candidate to
be used a data persistence technology for sensor networks.

\subsection{Advantages of Distributed Systems}

Distributed Systems can improve the performance of data management by providing
techniques for parallel data processing such as Map Reduce \cite{map-reduce,
bigtable}. It also helps decreasing the system load and bottleneck problems
related data creation and retrieval. It is also seen as an optional mechanism
of reducing the funneling affect of a centralized system by directing queries
to a given data partition \cite{sn-storage04}. Data Replication
\cite{data-replication} provides a ways to decentralize data access by
providing duplicated data that is intended to be used on reading operations.

\subsection{Disadvantages of Distributed Systems}

Requires more technical knowledge expertise to manage the different systems
that composed a distributed database.

@@

\begin{itemize}
  \item \textbf{Open-Source}: Given the nature of the project, the
  technology to be used must be free of charge \cite{open-source}, that is, no
  costs involved in the adoption a such technology; along with the support from
  the community;
  \item \textbf{Native APIs}: In order to consider the scope of this work
  defined in the last chapter, the solution for persisting collected sensor
  data must be not only limited to a technology that provides data access, such
  as SQL, but also by other data access mechanisms \textbf{data access
  mechanisms};
  \item \textbf{Export Feature}: the technology must support export
  capabilities.
\end{itemize}

@@

Most of the technologies listed provides access to the data sets through the
use of drivers in different programming and scripting languages such as Java,
Python, Perl, among others. Moreover, only TinyDB does not support hot backup.

>>>>

%\bibliography{thesis}  % uses the references stored in Chapter1Radar.bib
\setcounter{table}{1}
\chapter{Empirical Analysis of Data Persistence Taxonomies for NetBEAMS}

@@

\begin{itemize}
  \item Provide data management over the collected sensor data, as a
  conventional database systems;
  \item Must be able to scale in any increases to the data load of
  the sensor network, as described in Chapter 2;
  \item  Must be able to use single or multiple servers to improve performance
  and optionally deal with Data-Centric approaches to sensor data partitioning;
  \item Must provide a data model that scales, decreasing the constant data
  schema changes.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: an open-source relational database
  system popularized by the development of Internet applications, maintained
  by Sun Microsystems. \cite{sn-dataware-house} reports its use to develop a
  data warehouse for data collected from sensor devices using this technology;
  \item \textbf{TinyDB}\cite{tinydb}: developed and maintained by the Computer
  Science department of University of California, Berkeley, TinyDB is a
  relational database system specially developed for sensor devices. As
  referred in Section \ref{sec:sn-persitence-storage}, it is system used in
  many different sensor networks implementation such as \cite{sn-db-tinydb,
  sn-db-newop};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source KVP database
  system suggested by different surveys to be an alternative to relational
  databases \cite{db-is-rdbs-dommed, cloud-comp-architectures}. It is maintained
  by 10Gen Incorporated;
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. It is privately owned by IBM.
\end{itemize}

The following sections review these technologies' capabilities to adhere to
the specifications of the taxonomies defined in Chapter 3, as well as the
requirements and the scope for a data persistence of NetBEAMS, as defined in
Chapter 4.

@@

The advantages of the External Storage approach is the most common way to
implement persistence for any type of systems since data management is usually
offered by most database systems out-of-the-box. It is basically used to store
small volumes of data. On the other hand, the most common problems related to a
single External Data Storage approach are related to the capacity of the
device and operational problems. If the external storage device is situated in
the network, then the capacity of the device may be limited to receive
collected data from the sensors in the network, since it may run out of space.
Similarly, storage devices are very error prone and its hardware part may fail.

As shown in the previous chapter, NetBEAMS's architecture is based on sensor
network whose topology is a single-hop star with a centralized data sink. In
this way, all the collected data is transmitted to that single location as
described in Section \ref{sec:sn-infrastructure}. The simplest location to save
the collected data is through the use of at least an \textbf{External
Storage}. Similarly, the use of a Data-Centric Storage strategy is considered
in situations where the data sink is about to reach its storage capacity. The
\textbf{Data-Centric Storage} separates the collected data into dissimilar
locations, based on any property of the data. Therefore,
there are two different approaches to data partitioning:

\begin{itemize}
  \item \textbf{Horizontal Partitioning}: the data segmentation is implemented
  based on the values of the table rows, in the relational model point of view,
  or the collections elements, on KVP databases. That is, all the values defined
  by the columns are used in a given table physically located in specific
  servers. For instance, horizontal partitioning can be defined as a
  collection of five years worth of historical sensor data, partitioned into
  five separate servers. In this way, each server contains data related to
  each of the years;
  \item \textbf{Vertical Partitioning}: data is partitioned vertically in a way
  that the columns of an entity on the relational model are divided into
  different database tables. In this way, the particular dataset is broken
  down to different tables. For example, if the database is used to store
  pictures of a given sensor camera in a BLOB column, it would be the most
  preferable to the scenario where the images are not constantly referenced
  when compared to the textual data. Since KVP uses denormalized data, this
  partitioning strategy is not used.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Hash Partitioning}: a hash key is computed based on
  different table columns of an entity \cite{db-shard-schemas,
  db-partitioning-relational}, providing an even distribution of the data;
  \item \textbf{Key Partitioning}: is based on the hash partitioning over the
  values of a specific key \cite{db-mongo-partition};
  \item \textbf{Range Partitioning}: the partition is based on specific ranges
  of the values of given a given column of a table. For instance, the collected
  data from the sensors could be separated by the year of its collection. For
  example, data collected in the year of 2009 would be stored in a different
  partition than the data collected in the year 2010
  \cite{db-partitioning-relational}.
\end{itemize}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/database-sharding-by-region}
  \caption{Example of Data-Centric Storage for Sensor Networks using Database
  Sharding}
  \label{fig:database-sharding-by-region}
\end{figure}

@@

\begin{itemize}
  \item \textbf{Rebalancing Partitions}: considering a schema-dependent database
  system, changes on the schema of one shard requires rebalancing the
  collected data on each of the shards. As a consequence, the database shards
  transfers data to new locations whenever necessary. This technique is
  starting to appear as COTS implementations such as MySQL \cite{mysql} and
  mongoDB \cite{mongodb};
  \item \textbf{Referential integrity}: also considering a schema-dependent
  database system, any cross-shard query may run into the problem where
  references do not exist in other shards, since the application layer is
  responsible for enforcing data integrity. In this way, examples of such
  include verifying constraints of foreign keys when the partition is done
  using Vertical table partitioning, considering that data collections are
  organized in separate spaces, and physically placed in different shard nodes
  \cite{db-partitioning-relational}.
\end{itemize}

@@

One of the most common practices in the area of database system is to use the
relational model for data persistence, although the application of the system
may not fit to solve the problem. In fact, the main users of sensor networks may
not possess any expertise in database systems or data modeling, based on the
fact that they are from different science areas. \cite{sn-programming-language}
suggested the use of programming languages abstraction, instead of regular
database systems, to provide access to the collected data for both
technical and non-technical users. For instance, marine biologists manage
SF-BEAMS without expertise in data management or modeling systems. For this
reason, one of the requirements of the system is the use of a schema-less
approach, which describes the collected data without data modeling process
used in schema-dependent models. The \textbf{Key-Value-Pair Data Model} seems
to be an optional candidate that provides such a method of data design.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[scale=0.65]{../diagrams/Relational-Model-Original}
  \caption{Relational Data Model for NetBEAMS - A first prototype}
  \label{fig:Relational-Model-Original}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.6]{../diagrams/Relational-Model-Addition-Modified}
  \caption{Relational Data Model for NetBEAMS - Modified version with new
  entity}
  \label{fig:Relational-Model-Addition-Modified}
\end{figure}

\begin{itemize}
  \item Data schema that constantly change requires continuous database
  normalization processes, changes to structure, database management, and all
  the code that uses the data. Schema-Dependent databases always require
  refactoring of the schema to accommodate changes to the entities' properties;
  \item \cite{sn-provenance} describes the relational model as problematic 
  when describing collected data from sensor devices;
  \item \cite{sn-db-newop} suggests changes to the SQL clauses of TinyDB to
  better support time-series, an important property when describing
  collected data from with spatial-temporal properties.
\end{itemize}

Other projects have used the XML data models for persistence. It falls
into the same category as the Relational Model using XML documents \cite{xml},
since XML documents must fulfill the specifications defined by an XML Schema
\cite{xml-schema}. This model uses XPath \cite{xml-xpath} technology for
querying documents, although some hybrid \cite{db2} technologies may still use
SQL \cite{sql} for that matter. In this way, the view of an example of database
state for the defined previous schema can be seen in Figure
\ref{fig:persistence-example-relational}.

@@

This data model has been used to solve data models for different types of
application. One such a problem is the one to represent entities by a list of
properties with keys and relating values, called Key-Value data model
\cite{db-kvp-in-relational01, db-kvp-in-relational02}. The author of
\cite{db-kvp-in-relational01} suggests that this is used as a technique used
when the addition of number of properties of an entity constantly changes, and
thus, requires constant refactoring of the relational data model. Considering
this data model for sensor network devices, whose properties are lists of keys
and relating values, an ``one-fits-all'' attempt to use the KVP data model
using the relational data model is depicted in Figure
\ref{fig:KVP-on-Relational-Model}, and a given example of the state of a
database using this schema in Figure
\ref{fig:persistence-example-relational-kvp}.

@@

The growth of distributed systems and the Internet have enabled the development
of powerful database servers that can be organized in the context of
infrastructure and how to represent models. A powerful up-coming approach is
the use of database systems that do not use structured language for querying
its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, and it is characterized as follows:

@@

There are no records of the use of this data model with sensor networks. Data
is generally aggregated by given entities and any needed property applied to
it. Key may be composed by embedded keys having a dynamic structure and freely
used without a schema validation mechanism as used schema-dependent data
models, having the following organization:

\begin{itemize}
  \item entities from the same domains are placed into a ``bucket'';
  \item entities have a set of attributes and relating values;
  \item entities with different set of attributes may be contained in the same
  bucket, since there is no schema to govern the bucket items restriction.
\end{itemize}

For instance, all data needed for the YSI Sonde Data, as well as all necessary
provenance data and annotation as defined in Section
\ref{sec:sn-data-description}, are represented by means of key-value pairs as
seen in Figure \ref{fig:persistence-example-kvp}.

@@

The properties of the relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and summarized in Table 5.1:

\begin{table}[!b]
    \label{tab:properties-schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model is defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships \cite{db-normalization};
      \item Data Provenance can be provided through data types such as
      ``timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enables them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\newpage

In order to have access to the persisted data, both types of data models
support methods: one via structured languages, and the other via abstractions
of programming languages. \cite{db-is-rdbs-dommed} summarizes the differences
shown in Table 5.2:

\begin{table}[!h]
    \label{tab:daccess-schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic CRUD (Create-Retrieve-Update-Delete)
      database operations are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, containing aggregation functions to apply on the returned values.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via an abstraction of
      programming language. (E.g. ``db.collection.find()'',
      ``db.collection.insert()'');
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like equals to (=), different than (!=), less than
      or equals ($\leq$), greater than or equals ($\geq$), among others;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

Finally, each data model provides its own APIs to the integration with external
systems. \cite{db-is-rdbs-dommed} differentiates each of the application interfaces
provided by each of them, as summarized on table 5.3.

\begin{table}[!h]
    \label{tab:api-interface-schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC (Open Database
      Connectivity);
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Tend to provide Web Services Interfaces such as the REST
      \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

The data model used by each of the selected database systems can be summarized
as follows:

\begin{itemize}
  \item \textbf{MySQL}: uses the Relational Data Model with the use of SQL;
  \item \textbf{TinyDB}: uses the Relational Data Mode with the use of SQL; 
  \item \textbf{MongoDB}: uses Key-Value Pair Data Model with an abstraction of 
  programming language functions calls;
  \item \textbf{IBM DB2}: uses the Relational and Structured models with SQL
  and XPath, respectively.
\end{itemize}

@@

\subsection{Use of Provenance-Aware Data}

The inclusion of provenance data into the collected data enriches the
description of a given event, and is always useful to determine different
aspects of the data. As mentioned in Chapter 3, data provenance helps identify
the nature of the data by the identification of the properties of a given
sensor device, for example. Similarly, provenance adds spatial-time data in
order to correlate the observed data with time a given location and time of
occurrence.

Although metadata gives more information about the data, it just
represents additional data inserted together with the raw data collected
from the sensor devices may contribute with a decreasing amount of disk
storage. Therefore, the use of provenance-away data increases the required
disk space used to save the collected data from sensor devices.

\subsection{Use of Annotations}

Annotations are used to describe the collect data even more, since the
additional ``observation'' of the data are being stored together with the data.
For example, the use of tags to a given entity may help identifying and different
observations regarding the data collected from sensor devices. As shown in
Figure \ref{fig:persistence-example-kvp}, the addition of the key ``tag'' was
used to describe the observations that happened when the event of ``oil spill''
occurred. In this way, tags can be indexed for better search of the collected
data.

On the other hand, the use of annotations behaves the same as the use of
Provenance data. It will require additional disk space to store the
collected data from sensors.

@@

All the database systems technology supports the use of Data Provenance or
Annotations. The Key-Value Pair data model has the advantage of not maintaining
a schema design, and therefore, additional keys of descriptions can be added at
any time without requiring schema changes. On the other hand, the relational
data model requires schema changes to support the addition a new table column.

\begin{itemize}
  \item \textbf{MySQL}: supports data provenance and annotations via column
  types on the entities, requiring schema changes;
  \item \textbf{TinyDB}: supports data provenance and annotations via column
  types on the entities, requiring schema changes. Different projects have
  suggested changes to the SQL language used to give better support to
  spatial-temporal data \cite{sn-db-newop}.
  \item \textbf{MongoDB}: supports both data provenance and annotations with
  the definition of a new key on a given collection. No changes are required to
  the addition of the keys;
  \item \textbf{IBM DB2}: supports data provenance and annotations via column
  types on the entities, requiring schema changes.
\end{itemize}

@@

An advantage of centralized data management and query processing is simpler
than the in-network query method. Data is verified by a straightforward
database management system without the risk of data being unavailable, when
the data is spread among the network nodes. On the other hand, when the
Centralized Query processing is used, problems may occur. Depending on the
size of the datasets and the database technology, performance may be a concern
due to the alleged Funneling Affect \cite{sn-storage04}, since the
point-of-traffic is concentrated in the centralized system.

@@

Sensor Network data can be saved into database system that are organized
by either a Centralized or Distributed System. While a centralized database
system may be considered to be easier to manage, it may fail to adjust to
changes of requirements to manage data. On the other hand, a distributed
database system can be used to implement a data-centric solution. Therefore, a
database system that can be organized in both approaches is a good candidate to
be used a data persistence technology for sensor networks.

Distributed Systems can improve the performance of data management by providing
techniques for parallel data processing such as Map Reduce \cite{map-reduce,
bigtable}. It also helps decreasing the system load and bottleneck problems
related data creation and retrieval. It is also seen as an optional mechanism
of reducing the funneling effect of a centralized system by directing queries
to a given data partition \cite{sn-storage04}. Data Replication
\cite{data-replication} provides a way to decentralize data access by
providing duplicated data that is intended to be used on reading operations.

Although a distributed system may provide better support to the problem being
solved, it may be a difficult solution to propose. Since it is designed as set
of cooperating subsystems, multiple points of failure exist in this type of
system, making it difficult to be operated. Furthermore, it requires more
technical knowledge expertise to manage the different systems that composed a
distributed database management system.

@@

\begin{itemize}
  \item \textbf{Open-Source}: Given the nature of the project, the
  technology to be used must be free of charge \cite{open-source}, that is, no
  costs involved in the adoption a such technology; along with the support from
  the community;
  \item \textbf{Native APIs}: In order to consider the scope of this work
  defined in the last chapter, the solution for persisting collected sensor
  data must be not only limited to a technology that provides data access, such
  as SQL, but also by other data access mechanisms \textbf{data access
  mechanisms};
  \item \textbf{Export Feature}: the technology must support export
  capabilities during its execution such as hot backps.
\end{itemize}

@@

Most of the technologies listed provides access to the data sets through the
use of drivers in different programming and scripting languages such as Java,
Python, Perl, among others.

\begin{itemize}
  \item \textbf{MySQL}: Is  open-source, supports hot backup and contains
  scripting APIs from  the community;
  from the community. Also APIs are available for the majority of languages;
  \item \textbf{TinyDB}: Not open-source nor does it support hot backup;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from  the community, offering native APIs in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not open-source, but offers support from its
  community.
\end{itemize}
